---
title: "Ecological predictors of viral divergence"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Ecological predictors of viral divergence}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.2)
```

```{r setup}
library(viroscape)
library(ggplot2)
```

## The question

Influenza A evolves fast, but the rate varies by orders of magnitude between
places. Most models of that variation look inward: polymerase fidelity, host
immunity, transmission bottlenecks. This package is built around the outward
question instead. Does the ecological context a virus is circulating in predict
how far its sequences have drifted?

The response variable throughout is divergence from a regional consensus. The
predictors are country-year environmental and land-use indicators. The output is
a model that says which of them carry information once country identity and
sampling year are controlled for.

## The data

The bundled example pairs simulated H5-like haemagglutinin proteins with real
World Bank indicators for twelve countries. The sequences are generated *from*
those indicators with planted effects - agricultural land positive, livestock
index negative - so everything below recovers a construction. It exists to show
what the interface does, not to say anything about influenza. Reports built from
it carry a banner saying so.

```{r}
seqs <- example_sequences()
seqs
```

Host is recovered from the strain name, which is what makes it possible to
stratify by wild bird, domestic poultry and mammalian isolates.

```{r}
table(seqs$meta$host_class)
```

```{r}
env <- example_environment()
head(env[c("country", "year", "agricultural_land_pct", "livestock_index")])
```

## Alignment and the substitution model

```{r}
aln <- align_sequences(seqs, method = "none")   # already aligned
aln
```

Genetic distance depends on the substitution matrix assumed, so the matrix is
chosen rather than defaulted. Each candidate is fitted on the *same* fixed
neighbour-joining tree, so only the matrix varies between them, and they are
ranked by BIC.

```{r}
msel <- select_substitution_model(aln, criterion = "BIC")
msel$table
```

BIC rather than AIC here: its penalty scales with alignment length, so it is
the more conservative guard against an unnecessarily parameterised matrix.

```{r}
plot_model_selection(msel)
```

## Divergence from consensus

```{r}
cons <- build_consensus(aln)
cons
```

```{r}
dists <- distance_from_consensus(aln, cons, metric = "ml_aa", model = msel)
dists
```

No molecular clock correction is applied. The aim is regression on
environmental predictors rather than phylogenetic inference, and not assuming a
strict clock keeps the metric robust to rate variation across lineages.

Divergence from a consensus is emphatically *not* a rate, and it cannot be
turned into one by dividing by elapsed years: the consensus is not an ancestor
and the earliest cohort has no elapsed time to divide by. (Version 0.1.0 had a
`clock_correct` argument that did exactly that. On data simulated under a
perfectly constant rate it produced a strong, overwhelmingly significant
*negative* association with year - it manufactured the artefact its name
promised to remove. It has been removed and now raises an error.) If a rate is
what you want, root a tree and regress root-to-tip distance on collection date:

```{r}
rate <- divergence_rate(aln, by = "country", model = msel)
head(rate)
```

The slope is substitutions per site per year and the intercept implies a date
for the root. Tips are not independent, so treat the standard errors as
optimistic.

Nucleotide metrics are available too, and recover the synonymous change that
amino acid distances cannot see:

```{r}
distance_metrics()
```

## Joining the predictors

```{r}
ds <- join_environment(dists, env)
ds
```

Sampling is deliberately uneven, as it is in the real deposition record. This
is the plot to look at before trusting any country coefficient.

```{r}
plot_sampling(ds)
```

## The level the predictors live at

This is the step that decides whether anything downstream means what it says.
The response is measured per sequence. The predictors are country-year
quantities, so every sequence from one country-year carries an identical
predictor row, and those sequences are near-identical by descent as well.

```{r}
cluster_summary(ds)
```

The design effect is how many sequences it takes to carry the information of
one independent observation. Every criterion computed per sequence - AIC, the
GVIF, the likelihood ratio, `summary()`'s standard errors - is inflated by
roughly that factor. On simulated data with **no** environmental effect at all,
sequence-level forward selection retains at least one pure-noise predictor in
100% of runs, and a mean of four and a half predictors out of seven.

The fix is to model at the level the predictors vary at:

```{r}
mds <- aggregate_to_cluster(ds)
mds
```

## Structural terms first

Before any environmental predictor is considered, the country term has to be
settled.

```{r}
st <- select_structure(mds, criterion = "AIC")
st
```

Note that `year` is retained whether or not it improves the criterion.
Divergence from a consensus accumulates with sampling date, so year confounds
every predictor with a temporal trend: it earns its place by being a
confounder, not by fitting well. Set `force_time = FALSE` for the older
behaviour, which dropped it whenever it failed to improve AIC by 2.

A random intercept is the more natural specification, but its variance
component is only reliable with roughly ten or more groups. The example has
twelve, so it is a legitimate option here; the five-country design of the
original study was not. `select_structure()` warns when you are below that.

## Forward selection

```{r}
sel <- forward_select(mds, base_terms = st, criterion = "AIC", verbose = FALSE)
sel
```

Two things are worth reading out of that step log.

The first is why AIC is used inside a round rather than a likelihood ratio
test. Candidates within a round are not nested in one another, and the
likelihood ratio statistic only follows its chi-squared null distribution under
nesting.

The second is how collinearity is treated. A high GVIF that comes from
structural correlation with the country factor is a different problem from
redundancy between two predictors, so by default an exceedance is recorded but
does not reject a candidate on its own. What rejects a candidate is
destabilisation: a coefficient changing sign *and* losing significance, or a
significant non-grouping coefficient shifting by more than the threshold and by
more than one standard error.

```{r}
plot_selection_path(sel)
```

## Calibrating the search

Aggregating fixes the pseudoreplication. It does not fix the search. Seven
candidates tested at `delta_threshold = -2` is roughly seven tests at alpha
0.046, and on pure noise the cluster-level procedure still retains something in
about a third of runs. The only honest way to read a delta is against the
distribution of deltas the same search produces when the predictors are known
to be meaningless.

```{r}
cal <- calibrate_selection(mds, base_terms = st, criterion = "AIC", B = 199, seed = 1)
cal
```

Whole country-year predictor vectors are reassigned to other country-years, so
the clustering and the collinearity of the real data survive the permutation
and only the association with divergence is destroyed. Note how much larger
than 2 the honest 5% threshold is.

Attach it to the selection and it is reported alongside the step log, so a
delta is never shown without the null it should be read against:

```{r}
sel <- forward_select(mds, base_terms = st, criterion = "AIC",
                      calibration = cal, verbose = FALSE)
```

You do not have to remember to do this. `forward_select()` calibrates by
default, and when the calibrated threshold is stricter than `delta_threshold` it
is the one the search uses - so the honest threshold is applied during
selection, not offered afterwards to whoever reads the warning:

```{r}
forward_select(mds, base_terms = st, criterion = "AIC", verbose = TRUE)$thresholds
```

## Validation

The models along the selection path are nested, so a likelihood ratio statistic
can be computed - but the terms were chosen on the same data, by searching, so
the statistic does not have its chi-squared null distribution. The p-values are
anti-conservative and are labelled `nominal_p` to keep that visible. They
describe the fitted path; they are not evidence against a null hypothesis.

```{r}
lrt_chain(sel)
```

```{r}
gvif_table(sel$model)
```

```{r}
plot_gvif(sel)
```

```{r}
plot_coefficients(sel)
```

## Did it recover the truth?

The example data was generated with a known ground truth: agricultural land
positive, livestock index negative, plus a temporal trend. So the answer is
checkable.

```{r}
co <- tidy_coefficients(sel)
co[co$term %in% c("agricultural_land_pct", "livestock_index", "year"), ]
```

Two things about that table. The standard errors are cluster-robust, not
model-based. And the `unit` column is there because `join_environment()`
standardises predictors by default, so those coefficients are per standard
deviation of the predictor rather than per percentage point of agricultural
land - a distinction that is easy to lose between the analysis and the write-up.
`unscale()` converts:

```{r}
unscale(co)[co$term %in% c("agricultural_land_pct", "livestock_index"), ]
```

## The tree is a confounder too

Country-year clustering is not the only structure in the data. Sequences in the
same lineage share their history, so an environmental coefficient that survives
country, year and clustering may still be a lineage effect wearing a country's
clothes - lineages are not distributed evenly across countries, and divergence
from a consensus is exactly the kind of response that inherits phylogenetic
correlation.

`assign_clades()` cuts the tree into monophyletic groups, and clade can then be
used like any other grouping variable:

```{r}
tree <- build_tree(aln, model = msel)
clades <- assign_clades(aln, k = 6, tree = tree)
table(clades)
```

The bundled example has almost no lineage structure to find, because each
simulated sequence is drawn independently from a single ancestor. That is a
property of the example, not a limitation of the method: on data with real
lineages, `cluster_summary(add_clades(ds, clades, cluster = TRUE))` typically
reports a higher ICC than country-year does.

The first thing to do with a clade factor is put it in the model, either as a
base term or as a random effect:

```{r, eval = FALSE}
ds_cl <- add_clades(ds, clades)
forward_select(ds_cl, base_terms = c("country", "year", "clade"))
```

That is the primary tool, and it is the blunt one on purpose: a categorical
term absorbs discrete lineage structure exactly. On simulated data whose
divergence is driven entirely by six clades, with no environmental effect at
all, a confounded predictor is correctly returned as null (p = 0.19) once clade
is in the model.

`phylo_gls()` is the secondary check. It fits the same regression under a
covariance derived from the tree and prints each coefficient beside its
ordinary least squares counterpart *and* beside the clade fit:

```{r, eval = FALSE}
phylo_gls(ds, tree, terms = c("year", "agricultural_land_pct", "livestock_index"))
```

Read `shrinkage_pgls` before `p_pgls`. A coefficient that loses most of its
magnitude once the tree is in the model has been explained by descent whatever
its p-value says - and where the two corrections disagree, believe the clade
column. Brownian and Pagel structures spread covariance smoothly along branch
lengths, so they approximate block structure rather than absorbing it, and
discrete clades are block structure. `print()` says so when it happens.

## Where the confounding bites

Restricting to the five countries of the original study reproduces the problem
that paper had to reason around: country identity and country-level predictors
can no longer be separated.

```{r}
sea <- names(sea_countries())
seqs_sea <- filter_sequences(seqs, countries = sea)
aln_sea <- align_sequences(seqs_sea, method = "none")
d_sea <- distance_from_consensus(aln_sea, metric = "ml_aa", model = "JTT")
ds_sea <- join_environment(d_sea, env)
ds_sea
```

```{r}
sel_sea <- forward_select(aggregate_to_cluster(ds_sea),
                          base_terms = c("country", "year"), verbose = FALSE)
sel_sea$steps[c("round", "predictor", "delta", "gvif_scaled", "accepted")]
```

With five groups the GVIF on land-use predictors inflates sharply, and whatever
survives has to be read as an association rather than an independent effect.
Widening the panel is not a nicety; it is what makes the coefficients mean what
they appear to mean.

## Reproducibility

For a full run, use the `targets` pipeline rather than calling the functions by
hand. Every stage is a node, so changing the metric or a threshold reruns only
what depends on it.

```r
targets::tar_make()
targets::tar_visnetwork()
```

## Interactively

```r
launch_app()
```
